Equalizer Programming Guide\\ {\footnotesize\mdseries \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf} {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}

### 2007-09-03 Developer/ProgrammingGuide/paper.tex
Update Programming Guide, disable data transmission API, some example cleanups
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

\title{Equalizer Programming Guide}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.4, \today
}

@@

\thispagestyle{empty}
\begin{figure}[ht]
  \hfill
  \includegraphics[width=4cm]{logo.pdf}\hfill
  \includegraphics[width=4cm]{vmml.pdf}\hfill\vspace{-1em}\\
%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
\end{figure}
\vfill

@@

\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Version & \bf Date     & \bf Changes \\
    \hline
    0.4         & Aug 31, 2007 & added distributed objects\\
    0.3         & Aug 26, 2007 & added application and render client\\
    0.2         & Aug 20, 2007 & added main function\\
    0.1         & Aug 19, 2007 & outlined the basic concepts\\
    \hline \multicolumn{3}{c}{\small
      \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
      {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}\\
  \end{tabularx}}

@@

@@

@@

The application subclasses Equalizer objects and overrides virtual
functions to implement certain functionality, e.g., the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are in concept similar to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list of all task methods can be found on the
website\footnote{\htmladdnormallink{http://www.equalizergraphics.com/documents/design/taskMethods.html}{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

@@

\begin{wrapfigure}{r}{.5\textwidth}
  \includegraphics[width=.5\textwidth]{cave.pdf}{\caption{\label{fConfig}An
      example configuration}}
\end{wrapfigure}

@@

An \textsf{eq::Node} is the representation of a single computer in the
system. It is one operating system process of the render client. All
node task methods are executed from the main application thread.

@@

{\footnotesize\begin{lstlisting}
    // 2. Equalizer initialization
    NodeFactory nodeFactory;
    if( !eq::init( argc, argv, &nodeFactory ))
    {
        EQERROR << "Equalizer init failed" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
    // 3. initialization of local client node
    RefPtr< eqPly::Application > client = new eqPly::Application( initData );
    if( !client->initLocal( argc, argv ))
    {
        EQERROR << "Can't init client" << endl;
        eq::exit();
        return EXIT_FAILURE;
    }
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
int Application::run()
{
    // 1. connect to server
    RefPtr<eq::Server> server = new eq::Server;
    if( !connectServer( server ))
    {
        EQERROR << "Can't open server" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}

@@

    if( !config )
    {
        EQERROR << "No matching config on server" << endl;
        disconnectServer( server );
        return EXIT_FAILURE;
    }
\end{lstlisting}}

@@

    EQLOG( eq::LOG_CUSTOM ) << "Config init took " << clock.getTimef() << " ms"
                            << endl;
\end{lstlisting}}

@@

\begin{wrapfigure}{r}{.5\textwidth}
  \subfigure[]{\label{fSync}\includegraphics[width=.22\textwidth]{sync}}\hfil
  \subfigure[]{\label{fAsync}\includegraphics[width=.22\textwidth]{async}}%
  {\caption{\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames for which the
tasks are already queued in the node and pipe threads for
execution. This enables overlapped execution and minimizes idle
times. The first \textsf{latency Config::finishFrame} return
immediately, since they have no frame to synchronize
upon. Figure~\ref{fSyncAsync} shows the execution of (hypothetical)
rendering tasks without latency~(\ref{fSync}) and with a latency of
one~(\ref{fAsync}).

@@

    clock.reset();
    while( config->isRunning( ) && maxFrames-- )
    {
        config->startFrame();
        // config->renderData(...);
        config->finishFrame();
    }
    const uint32_t frame = config->finishAllFrames();
    const float    time  = clock.getTimef();
    EQLOG( eq::LOG_CUSTOM ) << "Rendering took " << time << " ms (" << frame
                            << " frames @ " << ( frame / time * 1000.f)
                            << " FPS)" << endl;
\end{lstlisting}}

@@

    // 6. cleanup and exit
    server->releaseConfig( config );
    if( !disconnectServer( server ))
        EQERROR << "Client::disconnectServer failed" << endl;
    server = 0;
    return EXIT_SUCCESS;
}
\end{lstlisting}}

@@

    // else execute client loops 'forever'
    while( true ) // TODO: implement SIGHUP handler to exit?
    {
        if( !eq::Client::clientLoop( ))
            return false;
        EQINFO << "One configuration run successfully executed" << endl;
    }
    return true;
}
\end{lstlisting}}

@@

    EQINFO << "New InitData instance" << endl;
}
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
        FrameData()
            {
                reset();
                setInstanceData( &data, sizeof( Data ));
                EQINFO << "New FrameData " << std::endl;
            }
\end{lstlisting}}

\if 0
\subsection{Config}

\subsubsection{Config Initialization}

The \textsf{eqPly::Config} class holds
{\footnotesize\begin{lstlisting}
bool Config::init()
{
    // init distributed objects
    _frameData.data.color = _initData.useColor();
    registerObject( &_frameData );
    _initData.setFrameDataID( _frameData.getID( ));

    registerObject( &_initData );
\end{lstlisting}}
\fi

>>>>

\title{Equalizer Programming Guide\\
  {\footnotesize\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.5, \today
}

@@

\thispagestyle{empty}
\begin{figure}[ht]
  \hfill
  \includegraphics[width=4cm]{images/logo.pdf}\hfill
  \includegraphics[width=4cm]{images/vmml.pdf}\hfill\vspace{-1em}\\
%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
\end{figure}
\vfill

@@

\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Version & \bf Date     & \bf Changes \\
    \hline
    0.5         & Sep 3, 2007  & added config and partly pipe\\
    0.4         & Aug 31, 2007 & added distributed objects\\
    0.3         & Aug 26, 2007 & added application and render client\\
    0.2         & Aug 20, 2007 & added main function\\
    0.1         & Aug 19, 2007 & outlined the basic concepts\\
    \hline
  \end{tabularx}}

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
All functionality of Equalizer is accessed through the Equalizer client
library, which implements all functionality discussed in this document.

@@

Methods called by the application always have the form
\textsf{verb[Noun]}, whereas methods called by Equalizer (`Task
Methods') have the form \textsf{nounVerb}.

@@

The application subclasses Equalizer objects and overrides virtual
functions to implement certain functionality, e.g., the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are in concept similar to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list can be found on the
website\footnote{\htmladdnormallink{http://www.equalizergraphics.com/documents/design/taskMethods.html}{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/cave.pdf}
  {\caption{\small\label{fConfig}An example configuration}}
\end{wrapfigure}

@@

An \textsf{eq::Node} is the representation of a single computer in the
system. It is one operating system process of the render client. All
node task methods are executed from the main application thread. Each
configuration might also use an application node, in which case the
application process is also used for rendering.

@@

{\footnotesize\begin{lstlisting}
    // 2. Equalizer initialization
    NodeFactory nodeFactory;
    if( !eq::init( argc, argv, &nodeFactory ))
    {
        EQERROR << "Equalizer init failed" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
    // 3. initialization of local client node
    RefPtr< eqPly::Application > client = new eqPly::Application( initData );
    if( !client->initLocal( argc, argv ))
    {
        EQERROR << "Can't init client" << endl;
        eq::exit();
        return EXIT_FAILURE;
    }
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
int Application::run()
{
    // 1. connect to server
    RefPtr<eq::Server> server = new eq::Server;
    if( !connectServer( server ))
    {
        EQERROR << "Can't open server" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}%>>

@@

    if( !config )
    {
        EQERROR << "No matching config on server" << endl;
        disconnectServer( server );
        return EXIT_FAILURE;
    }
\end{lstlisting}}%>>

@@

    EQLOG( eq::LOG_CUSTOM ) << "Config init took " << clock.getTimef() << " ms"
                            << endl;
\end{lstlisting}}%>>

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.18\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.18\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames for which the
tasks are already queued in the node and pipe threads for
execution. This enables overlapped execution and minimizes idle
times. The first \textsf{latency Config::finishFrame} return
immediately, since they have no frame to synchronize
upon. Figure~\ref{fSyncAsync} shows the execution of (hypothetical)
rendering tasks without latency~(\ref{fSync}) and with a latency of
one~(\ref{fAsync}).

@@

    clock.reset();
    while( config->isRunning( ) && maxFrames-- )
    {
        config->startFrame();
        // config->renderData(...);
        config->finishFrame();
    }
    const uint32_t frame = config->finishAllFrames();
    const float    time  = clock.getTimef();
    EQLOG( eq::LOG_CUSTOM ) << "Rendering took " << time << " ms (" << frame
                            << " frames @ " << ( frame / time * 1000.f)
                            << " FPS)" << endl;
\end{lstlisting}}%>>

@@

    // 6. cleanup and exit
    server->releaseConfig( config );
    if( !disconnectServer( server ))
        EQERROR << "Client::disconnectServer failed" << endl;
    server = 0;
    return EXIT_SUCCESS;
}
\end{lstlisting}}%>>

@@

    // else execute client loops 'forever'
    while( true ) // TODO: implement SIGHUP handler to exit?
    {
        if( !eq::Client::clientLoop( ))
            return false;
        EQINFO << "One configuration run successfully executed" << endl;
    }
    return true;
}
\end{lstlisting}}%>>

@@

    EQINFO << "New InitData instance" << endl;
}
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
        FrameData()
            {
                reset();
                setInstanceData( &data, sizeof( Data ));
                EQINFO << "New FrameData " << std::endl;
            }
\end{lstlisting}}%>>

\subsection{Config}

The configuration is driving the applications rendering, that is, it is
responsible for updating the data based on received events, requesting
new frames to be rendered and to provide the render clients with the
necessary data.

\subsubsection{Config Initialization and Exit}

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
execution is synchronized later.

The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This is necessary to allow a speedy startup of
the configuration on large-scale graphics clusters. On the other hand,
it means that initialization functions are called even if the parent's
initialization has failed.

The \textsf{eqPly::Config} class holds the master versions of the
initialization and frame data. Both objects are registered with the
\textsf{eq::Config}, which is the \textsf{eqNet::Session} used for
rendering. Equalizer takes care of the session setup and exit in
\textsf{Client::choose\-Config} and \textsf{Client::releaseConfig},
respectively.

The frame data is registered first, since its identifier is transmitted
using the initialization data, which is registered afterwards. The
identifier of the initialization data is transmitted to the render
client nodes using the \textsf{initID} parameter of
\textsf{eq::Config::init}. Equalizer will pass this identifier to all
\textsf{configInit} calls of the respective objects:

{\footnotesize\begin{lstlisting}
bool Config::init()
{
    // init distributed objects
    _frameData.data.color = _initData.useColor();
    registerObject( &_frameData );
    _initData.setFrameDataID( _frameData.getID( ));

    registerObject( &_initData );

    // init config
    _running = eq::Config::init( _initData.getID( ));
    if( !_running )
        return false;
\end{lstlisting}}

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking device, as this is an orthogonal
problem to parallel rendering, and has been solved already by a number
of implementations\footnote{VRCO trackd, OpenVPRN, etc.}. The example
code in \textsf{eqPly} is one reference implementation for the
integration of such a tracking library:

{\footnotesize\begin{lstlisting}
    // init tracker
    if( !_initData.getTrackerPort().empty( ))
    {
        if( !_tracker.init( _initData.getTrackerPort() ))
            EQWARN << "Failed to initialise tracker" << endl;
        else
        {
            // Set up position of tracking system in world space
            // Note: this depends on the physical installation.
            vmml::Matrix4f m( vmml::Matrix4f::IDENTITY );
            m.scale( 1.f, 1.f, -1.f );
            //m.x = .5;
            _tracker.setWorldToEmitter( m );

            m = vmml::Matrix4f::IDENTITY;
            m.rotateZ( -M_PI_2 );
            _tracker.setSensorToObject( m );
            EQLOG( eq::LOG_CUSTOM ) << "Tracker initialised" << endl;
        }
    }

    return true;
}
\end{lstlisting}}%>>

The exit of the configuration stops the render clients by calling
\textsf{eq::Config::exit}, and then deregisters the initialization and
frame data objects with the session:

{\footnotesize\begin{lstlisting}
bool Config::exit()
{
    _running = false;
    const bool ret = eq::Config::exit();

    _initData.setFrameDataID( EQ_ID_INVALID );
    deregisterObject( &_initData );
    deregisterObject( &_frameData );

    return ret;
}
\end{lstlisting}}

\subsubsection{Frame Control}

The rendering frames are issued by the application. The \textsf{Config}
only overrides \textsf{startFrame} in order to update the its data
before forwarding the start frame request to the \textsf{eq::Config}.

If a tracker is used, the current head position and orientation is
retrieved and given to Equalizer, which uses the head matrix together
with the wall or projection description to compute the view
frustra\footnote{see
  \htmladdnormallink{http://www.equalizergraphics.com/documents/design/immersive.html}{http://www.equalizergraphics.com/documents/design/immersive.html}}.

The camera position is updated and the frame data is commited, which
generates a new version. This version is passed to the rendering
callbacks and will be used to synchronize the frame data to the state
belonging to the current frame:

{\footnotesize\begin{lstlisting}
uint32_t Config::startFrame()
{
    // update head position
    if( _tracker.isRunning() )
    {
        _tracker.update();
        const vmml::Matrix4f& headMatrix = _tracker.getMatrix();
        setHeadMatrix( headMatrix );
    }

    // update database
    _frameData.data.rotation.preRotateX( -0.001f * _spinX );
    _frameData.data.rotation.preRotateY( -0.001f * _spinY );
    const uint32_t version = _frameData.commit();

    return eq::Config::startFrame( version );
}
\end{lstlisting}}

\subsubsection{Event Handling}

Events are send by the render clients to the application using
{eq::Config::sendEvent}. At the end of the frame,
\textsf{Config::finishFrame} calls \textsf{Config::handleEvents} to do
the event handling. The default implementation processes all pending
events by calling \textsf{Config::handleEvent} for each of them.

For event-driven execution, the application can override
\textsf{Config::handleEvents} to blockingly receive events using
\textsf{Config::nextEvent} until a new frame has to be rendered.

The \textsf{eqPly} example continuously renders new frames. It
implements \textsf{Config::hand\-le\-Event} to provide the various reactions
to user input:

{\footnotesize\begin{lstlisting}
bool Config::handleEvent( const eq::ConfigEvent* event )
{
    switch( event->type )
    {
        case eq::ConfigEvent::WINDOW_CLOSE:
            _running = false;
            return true;

        [...]

        default:
            break;
    }
    return eq::Config::handleEvent( event );
}
\end{lstlisting}}

\subsection{Node}

Foreach active render client, one \textsf{eq::Node} instance is
created on the appropriate machine. Nodes are only instantiated on their
render client processes, i.e., each process should have only one
instance of the \textsf{eq::Node} class. The application process might
also have a node class, which is handled in exactly the same way as the
render client nodes.

During node initialization the initialization data is mapped to
a local instance using the passed identifier from
\textsf{Config::init}. The model is loaded based on the filename in the
initialization data. No pipe, window or channel tasks methods are
executed before \textsf{Node::configInit} has returned.

{\footnotesize\begin{lstlisting}
bool Node::configInit( const uint32_t initID )
{
    eq::Config* config = getConfig();
    const bool  mapped = config->mapObject( &_initData, initID );
    EQASSERT( mapped );

    const string& filename = _initData.getFilename();
    EQINFO << "Loading model " << filename << endl;

    _model = PlyFileIO::read( filename.c_str( ));
    if( !_model)
        EQWARN << "Can't load model: " << filename << endl;

    return eq::Node::configInit( initID );
}
\end{lstlisting}}%>>

The node config exit deletes the loaded model and unmaps the
initialization data:

{\footnotesize\begin{lstlisting}
bool Node::configExit()
{
    delete _model;
    _model = NULL;

    eq::Config* config = getConfig();
    config->unmapObject( &_initData );

    return eq::Node::configExit();
}
\end{lstlisting}}

\subsubsection{Frame Control}

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/mainloop.pdf}
  {\caption{\small\label{fFrameSync}Synchronization of frame tasks}}
\end{wrapfigure}
The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entity
unlock all its children by calling \textsf{startFrame}, e.g., the
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls
are currently empty, as the synchronization is implicit due to the
threading model.

Likewise, \textsf{Config::finishFrame} causes the invokation of the
\textsf{frameFinish} task methods. These task methods unlock their
parents by calling \textsf{finishFrame}.

The explicit synchronization of child or parent resources allows the
application to optimize the processing, by doing certain, independent
operations when the child or parent resources are already unlocked.

\fig{fFrameSync} outlines the synchronization for the application, node
and pipe classes. The window and channel synchronization is similar and
omitted for simplicity. The \textsf{eqPly} example does not override
\textsf{Node::frameStart} or \textsf{frameFinish}, but it is absolutely
vital for the execution that \textsf{Node::startFrame} or
\textsf{Node::finishFrame} are called, respectively. The default
implementation of the node task methods does take care of that.

\subsection{Pipe}

All task methods for a pipe and its children are executed in a separate
thread. This approach optimizes usage of the GPU resource, since all
tasks are executed serially and do not compete for the usage, i.e.,
context switches are minimized. Later versions of Equalizer might
introduce threaded windows to allow the parallel and independent
execution of rendering tasks on a single pipe.

\subsubsection{Initialization and Exit}

Pipe threads are not explicitely synchronized with each other, that is,
pipes might be rendering different frames at one given time. Therefore
frame-specific data has to be allocated for each pipe thread, which in
the \textsf{eqPly} example is the frame data. The frame data is a member
variable of the \textsf{eqPly::Pipe}, and is mapped to the identifier
provided by the initialization data. The initialization in
\textsf{eq::Pipe} does GPU-specific initialization, e.g., the display
connection is opened when glX/X11 is used:

{\footnotesize\begin{lstlisting}
bool Pipe::configInit( const uint32_t initID )
{
    const Node*     node        = static_cast<Node*>( getNode( ));
    const InitData& initData    = node->getInitData();
    const uint32_t  frameDataID = initData.getFrameDataID();
    eq::Config*     config      = getConfig();

    const bool mapped = config->mapObject( &_frameData, frameDataID );
    EQASSERT( mapped );

    return eq::Pipe::configInit( initID );
}
\end{lstlisting}}

The config exit is again symmetric to the config initialization. The
frame data is unmapped and GPU-specific data is de-initialized by
\textsf{eq::Config::exit}:

{\footnotesize\begin{lstlisting}
bool Pipe::configExit()
{
    eq::Config* config = getConfig();
    config->unmapObject( &_frameData );

    return eq::Pipe::configExit();
}
\end{lstlisting}}

\subsubsection{Window System}

Equalizer supports multiple window system interfaces, at the moment
glX/X11, WGL and AGL/Carbon. Some operating systems, and therefore some
Equalizer versions, support multiple window systems\footnote{see
  \htmladdnormallink{http://www.equalizergraphics.com/compatibility.html}{http://www.equalizergraphics.com/compatibility.html}}.

Each pipe might use a different window system for rendering, which is
determined before \textsf{Pipe::configInit} by
\textsf{Pipe::selectWindowSystem}. The default implementation of
\textsf{selectWindowSystem} loops over all window systems and returns
the first supported window system, determined using
\textsf{supportsWindowSystem}.

The \textsf{eqPly} examples allows selecting the window system using a
command line option. Therefore the implementation of
\textsf{selectWindowSystem} is overwritten, and return the specified
window system, if it is valid:

{\footnotesize\begin{lstlisting}
eq::WindowSystem Pipe::selectWindowSystem() const
{
    const Node*            node     = static_cast<Node*>( getNode( ));
    const InitData&        initData = node->getInitData();
    const eq::WindowSystem ws       = initData.getWindowSystem();

    if( ws == eq::WINDOW_SYSTEM_NONE )
        return eq::Pipe::selectWindowSystem();
    if( !supportsWindowSystem( ws ))
    {
        EQWARN << "Window system " << ws 
               << " not supported, using default window system" << endl;
        return eq::Pipe::selectWindowSystem();
    }

    return ws;
}
\end{lstlisting}}%>>

\subsubsection{Frame Control}

TBD

{\footnotesize\begin{lstlisting}
void Pipe::frameStart( const uint32_t frameID, const uint32_t frameNumber )
{
    _frameData.sync( frameID );
    startFrame( frameNumber );
}
\end{lstlisting}}
